---
title: "Oxidative stress constrains the evolution of phage host-range diversity in a microbial community"
author: "Quentin Lamy-Besnier"
date: "`r Sys.Date()`"
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    theme: united
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Statistical analysis of the mutations identified in Meynard-Doumenc et al. 2024: Oxidative stress constrains the evolution of phage host-range diversity in a microbial community. 

See https://github.com/QLamyBesnier/Meynard-Doumenc_2024 for more information.

# Set-up

## Packages

Load the required packages:

```{r Packages, message=FALSE, warning=FALSE}
library(reshape2)
library(tidyr)
library(dplyr)
library(ResourceSelection)
```

## Data preparation

Load the necessary data:

```{r Data, message=FALSE, warning=FALSE}
# load the interaction matrix
interaction_matrix <- read.csv("interaction_matrix.csv", sep = ";", row.names = 1)

# load the list of mutations for each phage clone
phage_mutations <- read.csv("phage_mutations_matrix.csv", sep = ";")

# load the list of mutations for each bacterial clone
bacteria_mutations <- read.csv("bacteria_mutations_matrix.csv", sep = ";")
```

Preparation of the interaction matrix, which contains the infectivity each phage/bacteria couple (0 = no infection, 1 = infection):

```{r Interaction_matrix_preparation, message=FALSE, warning=FALSE}
# format data to long format
interaction_df <- as.data.frame(as.table(as.matrix(interaction_matrix)))
colnames(interaction_df) <- c("Phage", "Bacteria", "Interaction")

# remove NA from the interaction dataframe
interaction_df <- na.omit(interaction_df)
```

Preparation of the phage mutations data, which contains the list of mutations for each phage clone:

```{r Phage_mutations_preparation, message=FALSE, warning=FALSE}
# function to create binary matrices for the mutation data
create_binary_matrix <- function(mutation_df) {
  # format data to long format
  long_format <- melt(mutation_df, id.vars = NULL, variable.name = "Clone", value.name = "Gene")
  # remove missing values (if any)
  long_format <- long_format[!is.na(long_format$Gene), ]
  # create binary presence/absence matrix
  binary_matrix <- dcast(long_format, Gene ~ Clone, fun.aggregate = function(x) as.integer(length(x) > 0), fill = 0)
  # gene names as rownames
  rownames(binary_matrix) <- binary_matrix$Gene
  binary_matrix <- binary_matrix[ , -1]  # Remove the 'Gene' column
  return(t(binary_matrix))  # Transpose so that clones are rows and genes are columns
}

# create binary matrix for phage data
phage_gene_matrix <- as.data.frame(create_binary_matrix(phage_mutations))

# remove the useless "V1" and "Error" column (problematic clones will be processed later)
phage_gene_matrix <- subset(phage_gene_matrix, select = -c(V1, Error))

# add a "P" (for phage) prefix to all the phage genes so they can be distinguished 
colnames(phage_gene_matrix) <- paste0("P_", colnames(phage_gene_matrix))
```

Preparing the bacterial mutations data, which contains the list of mutations for each bacterial clone:

```{r Bacteria_mutations_preparation, message=FALSE, warning=FALSE}
# create binary matrix for bacteria data
bacteria_gene_matrix <- as.data.frame(create_binary_matrix(bacteria_mutations))

# remove the useless "V1" and "Error" column (problematic clones will be processed later)
bacteria_gene_matrix <- subset(bacteria_gene_matrix, select = -c(V1, Error))

# add a "B" (for bacteria) prefix to all the bacterial genes so they can be distinguished 
colnames(bacteria_gene_matrix) <- paste0("B_", colnames(bacteria_gene_matrix))
```

Merging all the information into a single dataframe.

```{r Final_dataset_interactions_mutations, message=FALSE, warning=FALSE}
# replace "." in the names for "-" for consistency
rownames(phage_gene_matrix) <- gsub("\\.", "-", rownames(phage_gene_matrix))
rownames(bacteria_gene_matrix) <- gsub("\\.", "-", rownames(bacteria_gene_matrix))
interaction_df$Bacteria <- gsub("\\.", "-", interaction_df$Bacteria)

# merge the gene matrices with the interaction data
interaction_df_phages <- merge(interaction_df, phage_gene_matrix, by.x = "Phage", by.y = "row.names")
interaction_df_final <- merge(interaction_df_phages, bacteria_gene_matrix, by.x = "Bacteria", by.y = "row.names")

head(interaction_df_final[,c(1:5)])
```

Exclusion of 5 clones for which we do not have Breseq information (MG-1-F2 and MG-4-F2 for bacteria; MEc-5-F1, LF-3-F2 and MG-3-F4 for phages):

```{r Data_cleaning, message=FALSE, warning=FALSE}
# remove the problematic clones
interaction_df_final_filtered <- interaction_df_final %>%
  filter(!(Bacteria %in% c("MG-1-F2", "MG-4-F2") | Phage %in% c("MEc-5-F1", "LF-3-F2", "MG-3-F4")))
```

Exclusion of 4 bacterial clones which have a lot of mutations:

```{r Data_cleaning_bis, message=FALSE, warning=FALSE}
# remove the clones with many mutations
interaction_df_final_filtered <- interaction_df_final_filtered %>%
  filter(!(Bacteria %in% c("MF-7-F1", "LF-2-F1", "LF-6-F1", "MEc-4-F2"))) 
```

# Analysis

We will look at each condition separately: H₂O₂ (1 fermenter) versus control (2 fermenters).

```{r Data_separation_condition, message=FALSE, warning=FALSE}
# data with only clones from the control fermenters (fermenters 1 and 2)
interaction_df_control <- interaction_df_final_filtered %>%
  filter(grepl("-F[12]$", Bacteria) & grepl("-F[12]$", Phage))

# data with only clones from the H₂O₂ fermenters (fermenter 4)
interaction_df_h2o2 <- interaction_df_final_filtered %>%
  filter(grepl("-F4$", Bacteria) & grepl("-F4$", Phage))
```

## Control

We start by the control condition and create a binomial model taking into account all the mutations:

```{r Control_model, message=FALSE, warning=FALSE}
# model creation
glm_int_ctrl <- glm(Interaction ~ ., data = interaction_df_control[, -c(1, 2)], family = binomial)

# model output
sum_glm_int_ctrl <- summary(glm_int_ctrl)
sum_glm_int_ctrl
```

Verification of the quality of the model:

```{r Control_model_verif, message=FALSE, warning=FALSE}
# mcfadden's r-squarred
1 - (glm_int_ctrl$deviance / glm_int_ctrl$null.deviance)

# hosmer-lemeshow test
hoslem.test(glm_int_ctrl$y, fitted(glm_int_ctrl), g=10)
```

Identification of the genes that have a statistically significant effect:

```{r Control_significant_all, message=FALSE, warning=FALSE}
# genes that have a significant effect
p_values_ctrl <- sum_glm_int_ctrl$coefficients[, "Pr(>|z|)"]

# Benjamini-Hochberg correction
p_values_adjusted_ctrl <- p.adjust(p_values_ctrl, method = "BH")

# significant genes after correction
names(p_values_adjusted_ctrl)[p_values_adjusted_ctrl < 0.05]

# full data for the significant genes after correction
sum_glm_int_ctrl$coefficients[names(p_values_adjusted_ctrl)[p_values_adjusted_ctrl < 0.05], , drop = FALSE]
```

4 genes are truly statistically significant, and they all favor phage infection.

## H₂O₂

Now, we look at the H₂O₂ condition (132 interactions). We make the exact same kind of model, taking into account all the possible mutations:

```{r H2O2_model, message=FALSE, warning=FALSE}
# model creation
glm_int_h2o2 <- glm(Interaction ~ ., data = interaction_df_h2o2[, -c(1, 2)], family = binomial)

# model output
sum_glm_int_h2o2 <- summary(glm_int_h2o2)
sum_glm_int_h2o2
```

Verification of the quality of the model:

```{r H2O2_model_verif, message=FALSE, warning=FALSE}
# mcfadden's r-squarred
1 - (glm_int_h2o2$deviance / glm_int_h2o2$null.deviance)

# hosmer-lemeshow test
hoslem.test(glm_int_h2o2$y, fitted(glm_int_h2o2), g=10)
```

Identification of the genes that have a statistically significant effect:

```{r H2O2_significant_all, message=FALSE, warning=FALSE}
# genes that have a significant effect
p_values_h2o2 <- sum_glm_int_h2o2$coefficients[, "Pr(>|z|)"]

# Benjamini-Hochberg correction
p_values_adjusted_h2o2 <- p.adjust(p_values_h2o2, method = "BH")

# significant genes after correction
names(p_values_adjusted_h2o2)[p_values_adjusted_h2o2 < 0.05]
```

No genes remain statistically significant after the multi-testing correction, which is expected as both were barely significant (p = 0.04 for both).





